Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa

### 2022-05-06 src/docs_12-05-22-mrc-gida/12-05-22-mrc-gida.Rmd
Presentable version of slides incase laptop dies again. Close #112

---
title: Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa
subtitle: MRC Centre for Global Infectious Disease Analysis Seminars
author: Adam Howes
institute: Imperial College London
date: May 2022
bibliography: citations.bib
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

@@

* In sub-Saharan Africa, adolescent girls and young women (AGYW) are XX.X\% of the population but XX.X\% of all new HIV infections
* Reasons for this disparity include:
  1. Younger age at first sex
  2. Age patterns of sexual mixing
  3. Structural vulnerabilities and power imbalances
  4. Increased susceptibility to HIV infection

@@

* To have the greatest impact with available prevention resources, interventions should be prioritised to those at highest risk
* The Global AIDS Strategy 2021-2016 proposed stratifying HIV prevention for AGYW based upon
  1. Population-level HIV incidence in the local population
  2. Individual-level sexual risk behaviour

@@

1. Estimate the proportion of AGYW in four sexual risk behaviour groups at a district level (in 13 countries identified as priority by The Global Fund)
2. Analyze numbers of new infections reached by stratified prevention strategies

@@

* Sexual behaviour data from AIS, BAIS, DHS and PHIA household surveys (see next slide)
* Later, we used district-level HIV incidence, prevalence, population size estimates from the Naomi model [@eaton2021naomi]

@@

* Multinomial logistic regression model for the proportion of AGYW in the none, low and {high, very high} risk groups, using all 47 surveys
* Selected model included:
  * Age effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Correlated temporal effects (AR1) 
* Multinomial-Poisson transformation allowed us to use `R-INLA` for inference

@@

* Logistic regression model for the proportion of those in the {high, very high} risk groups who are in the very high risk group, using only the 13 surveys with a specific transactional sex question
* Selected model included:
  * Age effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Clients of FSW covariates [@hodgins2022population]
* Also used `R-INLA` for inference here

@@

* Take 1000 samples from each model, then multiply suitably to generate estimates for all four risk groups
* We adjusted the very high risk group to match national FSW estimates from @johnston2022deriving

@@

```{r echo=FALSE, out.width = "400px"}
knitr::include_graphics("depends/aids-abstract-A.pdf")
```

@@

```{r echo=FALSE, out.width = "400px"}
knitr::include_graphics("depends/aids-abstract-B.pdf")
```

<!-- ## -->

<!-- ```{r echo=FALSE, out.width = "400px"} -->
<!-- knitr::include_graphics("depends/3p1-continental-map.png") -->
<!-- ``` -->

<!-- ## -->

<!-- ```{r echo=FALSE, out.width = "400px"} -->
<!-- knitr::include_graphics("depends/3p1-within-between-country-variation.png") -->
<!-- ``` -->

@@

### Integration of all relevant surveys

* Modular approach allowed estimating FSW proportion even for surveys without a specific transactional sex question

### Alleivating small-sample sizes

* Direct estimates display large variance
* We borrowed information across space, between countries and over surveys
* Our estimates more plausibly reflect spatial heterogeneity

@@

```{r echo=FALSE, out.width = "400px"}
knitr::include_graphics("depends/model-direct-benefits.png")
```

### Estimates even where there isn't direct data

* Although some people think of this as "making up data", the data almost never "speak for themselves"
* Uncertainty should be higher in regions with infilling: important to clearly communicate this

@@

* We hope that our risk group estimates will help countries to implement the Global AIDS Strategy
  * (User guide currently under development)
* It's very important to reach FSW!
* Countries have different epidemic profiles

@@

* `R-INLA` and the multinomial-Poisson trick are a relatively good choice for spatio-temporal multinomial models
  * Make good use of the `group` and `replicate` feature to define sufficiently complex Kronecker product effects
  * Can run into limitations on model complexity: I might be moving to something more flexible e.g. `tmb`, `aghq` to continue development of this project

@@

* \faEnvelope $\,$ `ath19@ic.ac.uk`
* \faHome $\,$ `athowes.github.io`

## Bibliography {-}

>>>>

---
title: Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa
subtitle: MRC Centre for Global Infectious Disease Analysis Seminars
author: Adam Howes
institute: Imperial College London
date: May 2022
bibliography: citations.bib
header-includes:
  - \widowpenalties 1 150
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

@@

* In sub-Saharan Africa, adolescent girls and young women (AGYW) aged 15-29 are disproportionately at risk of HIV infection
* This disparity is because of:
  1. Younger age at first sex
  2. Age patterns of sexual mixing
  3. Structural vulnerabilities and power imbalances
  4. Increased susceptibility to HIV infection

@@

* Prevention packages are made up of a core package, available to everyone, and more intensified interventions
* To have the greatest impact, it's important to priorisite intensified interventions to those at highest risk
* The Global AIDS Strategy 2021-2016 proposed stratifying HIV prevention for AGYW based upon
  1. Population-level HIV incidence in the local population
  2. Individual-level sexual risk behaviour

@@

1. Estimate the proportion of AGYW in four behavioural risk groups at a district level (in 13 countries identified as priority by The Global Fund)
2. Analyze numbers of new infections reached by stratified prevention strategies

@@

* We used sexual behaviour data from AIS, BAIS, DHS and PHIA household surveys to place respondents into four risk groups:
  1. $k = 1$: Not sexually active
  2. $k = 2$: One cohabiting sexual partner
  3. $k = 3$: Non-regular sexual partner(s)
  4. $k = 4$: Female sex workers
* District-level HIV incidence, prevalence, population size estimates from the Naomi model [@eaton2021naomi]
* Risk ratios from ALPHA network analysis [@slaymaker2020croi] and UNAIDS analysis led by Keith Sabin

```{r echo=FALSE, out.width = "400px"}
knitr::include_graphics("depends/category-flowchart.pdf")
```

@@

* Multinomial logistic regression model for the proportion of AGYW in the $k = 1, 2, 3^{+}$ risk groups, using all 47 surveys
* Selected model included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Correlated temporal effects (AR1) 
* Multinomial-Poisson transformation allowed us to use `R-INLA` for inference

@@

* Logistic regression model for the proportion of those in the $k = 3^{+} = \{3, 4\}$ risk groups who are in the $k = 4$ risk group, using only the 13 surveys with a specific transactional sex question
* Selected model included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Clients of FSW covariates [@hodgins2022population]
* Also used `R-INLA` for inference here

@@

* Take 1000 samples from each model, then multiply suitably to generate estimates for all four risk groups
* We adjusted the $k = 4$ risk group to match national FSW estimates from @johnston2022deriving

@@

```{r echo=FALSE, out.width="350px", fig.cap = "We found a geographic discontinuity in behaviour between Southern and Eastern Africa."}
knitr::include_graphics("depends/aids-abstract-A.pdf")
```

@@

```{r echo=FALSE, out.width="350px", fig.cap = "Here is another view of the discontinuity."}
knitr::include_graphics("depends/aids-abstract-B.pdf")
```

@@

* Integration of all relevant surveys
  * Modular approach allowed estimating FSW proportion even for surveys without a specific transactional sex question
* Alleivating small-sample sizes (that caused direct estimates to display large variance)
  * We borrowed information across space, between countries and over surveys so that our estimates more plausibly reflect spatial heterogeneity
* Estimates where there isn't direct data
  * Although some people think of this as "making up data", the data almost never "speak for themselves" (everything is a model)
  * Uncertainty should be higher in regions with infilling: important to transparently communicate this

@@

```{r echo=FALSE, out.width = "350px", fig.cap = "Illustration of the problem with direct survey estimates."}
knitr::include_graphics("depends/model-direct-benefits.png")
```

@@

* Our risk group estimates can be used to help countries implement the Global AIDS Strategy (user guide currently under development)
* It's important to reach FSW (not a new message!)
* Countries have different epidemic profiles

@@

* `R-INLA` and the multinomial-Poisson trick are a good choice for spatio-temporal multinomial models
  * The `group` and `replicate` features can be used to define quite complex Kronecker product effects
  * However, can run into limitations on model complexity and I might be moving to something more flexible e.g. `tmb`, `aghq` to continue development of this project

@@

* This project was joint work with members of the HIV inference group, particularly Katie Risher and Jeff Eaton
  * \faHome $\,$ `hiv-inference.org`
* You can find me via
  * \faEnvelope $\,$ `ath19@ic.ac.uk`
  * \faHome $\,$ `athowes.github.io`

## Bibliography {.allowframebreaks}
